Title

### 2012-01-17 inc/laser/laser.tex
All main text now ported from oe paper. Still need references, figures (and captions).

@@

>>>>

% figure 5

@@

The ML output from the femtosecond Yb:KGW laser with a net negative GVD of $-5,200\text{fs}^2$ per round-trip was frequency doubled in a critically phase-matched, 2mm, Type I Lithium triborate (LBO) crystal.
The crystal is uncoated and cut for frequency doubling with the fundamental $p$-polarized radiation at 1046nm incident at Brewster’s angle --- a geometry that ensures both minimum loss for the fundamental radiation and, more importantly, a high damage threshold.
The doubling crystal is positioned at the focus of a $z$-fold comprising a 1040nm high reflector with a radius of curvature of 75mm followed by a collimating mirror with a radius of curvature of 100mm reflective at both the fundamental and second harmonic wavelengths.
At the 3.5W ML output power, with a pulse duration of 250fs, the estimated focal irradiance for the fundamental is $\sim50\text{GW}/\text{cm}^2$ in the LBO doubling crystal.
Two plane dispersion compensation mirrors, each with a GVD of $-220\text{fs}^2$ (Layertec GmbH [32]), inserted before the $z$-fold doubling geometry compensate for propagation through the 6mm-thick fused silica output coupler, reflections off standard optics, etc.
After the collimating
%page 8639
mirror, a standard flat dichroic beamsplitter then separates the reflected $s$-polarized 523nm radiation from the transmitted $p$-polarized fundamental.

This frequency doubling geometry for LBO is chosen to be close to the optimum for efficient short pulse doubling [39,40] defined by limits imposed by group-velocity walk-off between the fundamental and second harmonic pulses, spatial walk-off in the critical phase-matching, and the ratio of the confocal parameter of the focused fundamental wave to the LBO crystal length.
In addition, the $z$-fold configuration allows for astigmatism compensation (e.g., due to the Brewster LBO crystal) of the second harmonic output by adjustment of the angle of incidence on the second collimating mirror.

Figure 5 shows the resultant frequency-doubled green power for the ML Yb:KGW laser.
The measured average power at 523nm increases rapidly from 0.5W at the 23A mode-locking threshold to over 1.3W at a diode drive current of 27A.
At and above 30A, the second harmonic power saturates at 1.65W --- an effect that is clearly consistent with the observed emergence of a DC spectral component at diode currents above 29A since only the ML pulse will be efficiently frequency doubled.
Also plotted in Fig. 5 is the frequency doubling efficiency, which is calculated taking into account the 17\% loss suffered by the $s$-polarized 523nm radiation upon exiting the Brewster-cut LBO crystal and the 4\% loss (to diagnostics) of the fundamental radiation prior to the $z$-fold doubling geometry.
The maximum doubling efficiency of over 60\% occurs at 27-28A of diode drive current when the ML laser output power is $\sim$3W.
At a ML output power of 3.5W (29-30A), the frequency doubling efficiency is still $\sim$60\% and slowly decreases above 30A as the DC spectral component absorbs the residual power above the soliton power limit.

\subsection{Summary}

A robust, high-power ($\sim$5W), diode-pumped and thermal-lens-shaped, femtosecond (220-350fs) Yb:KGW oscillator is presented.
The ML operation of the laser is self-starting above a critical output power of $\sim$2W, being initiated by a 1\% modulation depth SBR and subsequently supported in the soliton regime through a balance of self-phase modulation (Kerr lensing) in the gain medium and negative GVD (introduced using dispersion compensating mirrors).
The use of the TLS technique in the diode-pumped laser head provides for astigmatism-compensated TEM$_{00}$-mode operation (a spatial beam quality $M^2 <1.2$), while proper cavity design accommodates the strong Kerr lens (up to $\sim4\text{m}^{-1}$) in the Yb:KGW crystal under ML operation.
A full analysis of the thermal conduction in our 3x3x10mm Yb:KGW crystal is used to determine the correct TLS pump conditions, which lead to both excellent pump-laser beam overlap for the near collinear pump geometry and a near perfect pump-induced thermal lens (i.e., parabolic temperature distribution) experienced by the intracavity laser beam in the gain medium.

The laser head design also employs a simple single layer \ce{SiO2} anti-reflection coating on the gain medium to provide simultaneously (i) minimal loss for the intracavity 1040nm laser radiation $p$-polarized along the optical $N_p$-axis (crystallographic $b$-axis) associated with the largest gain bandwidth and (ii) less than 0.5\% surface reflection for the orthogonally $s$-polarized 980nm laser diode radiation (polarized parallel to the crystallographic $a$-axis) to enable efficient pumping of the strongest absorption line along the optical $N_m$-axis (rotated by $\sim20^\circ$ from the crystal $a$-axis).
In this near optimal pump-emission Yb:KGW crystal geometry for ML operation, the pump-induced thermal lens in the gain medium is observed to be positive, indicating that the sum of the thermo-optical coefficient and the stress-generated refractive index change with temperature is positive for 1$\mu$m radiation polarized parallel to the optical $N_p$-axis (crystallographic $b$-axis).

The 250fs ML pulses generated at 3.5W of average output power and a 63.5MHz repetition rate are frequency doubled in a 2mm-thick, Brewster-cut and critically phase-matched Lithium triborate (LBO) crystal.
The resultant efficient frequency doubling, which exceeds 60\% at a ML fundamental power of $\sim$3W, yields 1.65W of average power at 523nm ---
%page 8640
a peak green pulse power of $\sim$130kW.
This femtosecond visible output will be used as the driving pulsed laser radiation source for an ultrafast electron microscope [23].

%% Acknowledgements
% This research was supported in part by the National Science Foundation under award DMR-
% 0619573 and the Department of Energy under award DE-FG52-06NA26213. J.A.B.
% acknowledges support from the Department of Education under the Graduate Assistantships
% in Areas of National Need (GAANN) program.
